A Kochen-Specker system has at least 21 vertices\\ {\small (extended abstract)}

### 2014-05-11 main.tex
Fix biconnected and add a reference for the public lectures

The KS system proposed by Kochen and Specker contained 117 points\cite{ks}.
Penrose and Peres\cite{peres} independently found a smaller system of 33 points.
The current record is the 31 point system of Conway\cite[p.~197]{qtcm}.
As pointed out by \cite{c00,aow11}, finding small KS systems
is of both theoretical and practical interest.
In public lectures, Conway himself, stressed the search for small KS systems.

@@

We continue with a strengthening of Proposition~\ref{prop:ks-conn}.
\begin{prop}
A minimal Kochen-Specker graph is biconnected,
that is: removing any single edge leaves the graph connected.
\end{prop}
\begin{proof}
Suppose we have a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.  Thus~$B(G)$ is a tree.

Now consider a leaf~$A$ in the biconnected component graph~$B(G)$.
That is, $A$ is a biconnected component which is
only connected to one other biconnected component.
Let~$B$ be the union of the remaining biconnected components.
There is exactly one pair of vertices~$(a,b)$
such that~$a \in A$, $b \in B$ and $a$ is adjacent to~$b$.

Take any~$b_2 \in B$ with~$b_2 \neq b$.
By assumption~$b_2$ is not adjacent to~$a$.
Define~$G'$ to be the graph~$G$ extended with the edge~$(a,b_2)$.
Clearly, $G'$ is not~010-colorable.  We will show that~$G'$ is embeddable.

Consider an embedding of~$G$.
Observe that a rotation of all (points corresponding to) elements of~$A$
along the axis spanned
by (the point corresponding to)~$b$, preserves all relevant
orthogonality relations.

Rotate the vectors of the embedding
along~$b$ such that~$b_2$ becomes orthogonal to~$a$.
It could be that there are elements~$a'$ in~$A$
and~$b'$ in~$B$ that are rotated onto the same point.
Then consider rotations of all elements of~$B$ along~$a$.
There are finitely many angles for which two elements will overlap.
Hence, there is a rotation of all elements of~$B$ along~$a$
such that~$A$ and~$B$ do not overlap.
Note that all existing orthogonality relations are preserved.

These two rotations yield a new set of points on the sphere,
that demonstrate~$G'$ is embeddable.
Therefore, in generating all relevant graphs, we find both $G$ and
$G'$. Since they have the same order, we might as well ignore the
graph $G$. This shows it is sufficient to consider only biconnected
graphs.
\end{proof}
We believe a minimal KS graph is also triconnected.
However, we did not find a proof.

>>>>

The KS system proposed by Kochen and Specker contained 117 points\cite{ks}.
Penrose and Peres\cite{peres} independently found a smaller system of 33 points.
The current record is the 31 point system of Conway\cite[p.~197]{qtcm}.
As pointed out by \cite{c00,aow11}, finding small KS systems
is of both theoretical and practical interest.
In public lectures, Conway himself, stressed the search for small KS
systems.\cite{OC}

@@

We continue with a strengthening of Proposition~\ref{prop:ks-conn}.
\begin{prop}\label{prop:ks-biconn}
A minimal Kochen-Specker graph is biconnected,
that is: removing any single edge leaves the graph connected.
\end{prop}
We need some preparation, before we can prove this Proposition.
\begin{dfn}
Given a graph~$G$ and a vertex~$v$ of~$G$.
We say, \keyword{$v$ has fixed color~$c$ (in~$G$)},
if~$G$ is~$010$-colorable
and for every~$010$-coloring of~$G$,
the vertex~$v$ is assigned color~$c$.
\end{dfn}
We are interested in these graphs because of the following observation.
\begin{lem}
If there is an embeddable graph~$G$ on~$n$ vertices with a vertex
with fixed color~$1$,
then there is a Kochen-Specker graph on~$2n$ vertices.
\end{lem}
\begin{proof}
Let~$G$ be a graph and~$v$ a vertex of~$G$ with fixed color~$1$.
Consider two copies of the graph~$G$.
Connect the two instances of~$v$ with an edge.
Call this graph~$G'$.
Clearly, $G'$ is not~$010$-colorable.

We need to show~$G'$ is embeddable.
Given an embedding~$S$ of~$G$.
We may assume that the point in~$S$ corresponding to~$v$
is the north pole.
Furthermore, we may assume that there is no point on the $x$-axis,
by rotating points along the north pole.
Let~$S'$ be~$S$ rotated~$90$ degrees along the~$y$-axis.
Some points of~$S$ and~$S'$ might overlap.
That is: there might be a point~$s$ in~$S$ and~$s'$ in~$S'$
that are equal or antipodal.
Observe that if no points of~$S'$ and~$S$ overlap,
then~$S \cup S'$ is an embedding of~$G'$.

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlaping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.
\begin{comp}\label{comp:bic1}
Let~$F^{3,1}_n$ denote the number of connected graphs
with minimal vertex-order~$3$
on~$n$ vertices
with a vertex with fixed color~$1$.
Then:
\begin{center}
\begin{tabular}{l|lll}
    $n$ & $\leq 13$
        & $14$
        & $15$ \\
    \hline
    $F^{3,1}_n$ & $0$
        & $4$
        & $59$
\end{tabular}
\end{center}
Using the methods of Section~\ref{sec:emb},
we have determined these graphs unembeddable.
\end{comp}
We need one final computational result.
\begin{comp}\label{comp:bic2}
There are no connected graphs
on less than~$15$ vertices
with minimal vertex-order~$2$
with a vertex with fixed color~$0$. 
\end{comp}
We are ready to prove that a minimal Kochen-Specker graph
is biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Suppose we are given a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.  Thus~$B(G)$ is a tree.

Furthermore suppose~$G$ is a minimal Kochen-Specker graph.
Now consider a leaf~$A$ in the biconnected component graph~$B(G)$.
That is, $A$ is a biconnected component that is connected to only
one other biconnected component.
Let~$B$ be the remainder of the graph.
There is exactly one pair~$(a,b)$ with~$a$ in~$A$ and~$b$ in~$B$
and~$a$ adjacent to~$b$.

Note that~$A$ and~$B$ are~$010$-colorable.
Suppose that~$a$ does not have fixed color~$1$ in~$A$.
Then there is a~$010$-coloring of~$A$
which assigns~$0$ to~$a$.
But then~$G$ is~$010$-colorable, quod non.
Thus~$A$ is an embeddable graph with a vertex with fixed color~$1$.
Similarly~$B$ is an embeddable graph with a vertex with fixed color~$1$.
We may assume~$G$ is two copies of~$A$ connected at the
copies of the vertex with fixed color~$1$.

The graph~$G$ has minimal vertex-order~$3$.
Thus every vertex of~$A$ has minimal order~$3$ except for~$a$,
which has minimal order~$2$.
Suppose~$a$ has minimal order~$3$ in~$A$.
Then by Computation~\ref{comp:bic1},
the graph~$A$ has at least~$16$ vertices.
Thus~$G$ has at least~$32$ vertices, quod non.

Thus~$a$ must have order~$2$ in~$A$.
Let~$c,d \in A$ be the vertices connected to~$a$ in~$A$.
For~$a$ to have fixed color~$1$ in~$A$,
the vertices~$c$ and~$d$ must have fixed color~$0$ in~$A$.
By Computation~\ref{comp:bic2},
the graph~$A$ has at least~$15+1$ vertices.
Thus~$G$ has at least~$32$ vertices. Contradiction.
\end{proof}
We believe a minimal KS graph is also triconnected.
However, we did not find a proof.
